Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-21 doc/mc-paper.tex
more theta/phi corrections

This paper introduces \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
be added to $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
from $\phi \cup \{C\}$ to create $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following \textsc{final conditions} must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
any total assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

>>>>

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\theta$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\theta$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\theta$. We write $\theta_\text{foo}$ for
clausal encodings possibly using extension variables, and $\phi_\text{bar}$ for formulas that do not.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
from $\theta \cup \{C\}$ to create $\theta$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\theta$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following \textsc{final conditions} must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $\{r\}$ and the defining clauses of the POG nodes. Recognizing that
any total assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
falsify $\phi_\noder$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.
